---
title: "Group_project_clive's_work.Rmd"
author: "Clive Smith"
date: "2024-03-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#install.packages("sf")
```

```{r}
library(tidyverse)
library(lubridate)
library(janitor)
library(shiny)
library(shinydashboard)
library(sf)
```


```{r}
hospitals <- read_csv("data/healthcare_dataset.csv")
```

```{r}
hospitals <- clean_names(hospitals)
names(hospitals)
```
Here we removed the infromation that is not neeeded for the project. 
```{r}
hospitals <- hospitals %>%
  select(-blood_type, -room_number)
```

```{r}
hospitals_final <- hospitals %>%
  mutate(id = as.integer(factor(name))) %>%
  select(-doctor, -name) %>%
  arrange(id)
```

\\ Part 1

```{r}
medicare_cost <- hospitals[ , c("date_of_admission", "billing_amount")]
```

```{r}
pre_covid <- subset(medicare_cost, date_of_admission < as.Date("2020-01-01"))
post_covid <- subset(medicare_cost, date_of_admission >= as.Date("2020-01-01"))
summary(pre_covid$billing_amount)
summary(post_covid$billing_amount)
```


```{r}
combined_data <- bind_rows(pre_covid %>% mutate(period = "Pre-COVID"),
                           post_covid %>% mutate(period = "Post-COVID"))

# Plot combined data with separate lines of regression for pre and post-COVID data
combined_data %>%
  ggplot(aes(x = date_of_admission, y = log10(billing_amount), color = period)) +
  geom_point(size = 0.5) + 
  geom_smooth(method = "lm", aes(group = period), se = TRUE) +
  
  labs(title = "Medicare Costs Comparison",
       x = "Date of Admission",
       y = "Log10(Billing Amount)",
       color = "Period") +
  theme_minimal() +
  theme(plot.title = element_text(size = rel(1.75), hjust = 0.5))
```


part 2
#create a new column with price categories.
#create a graph and list for 3 things: price by hospitals, prices by insurance companies, and prices by condition.

```{r}
hospitals_final %>% 
  select(billing_amount) %>% 
  summary()
```

```{r}
hospitals_final %>% 
mutate(price_level=case_when(billing_amount<13507 ~ "low", 
                                  billing_amount>=13507 & billing_amount<=25258 ~ "normal", 
                                  billing_amount>25258 & billing_amount<=37734 ~ "high", 
                                  billing_amount>37734 ~ "very high"))
```
#this graph isn't feasible, there are just too many hospitals. We need to find a way to cut down on the number shown here somehow.
```{r}
hospitals_final %>% 
mutate(price_level=case_when(billing_amount<13507 ~ "low", 
                                  billing_amount>=13507 & billing_amount<=25258 ~ "normal", 
                                  billing_amount>25258 & billing_amount<=37734 ~ "high", 
                                  billing_amount>37734 ~ "very high")) %>% 
  group_by(hospital) %>% 
  count(hospital, price_level) %>% 
  ggplot(aes(x=hospital, y=n, fill=price_level))+
  geom_col(position="dodge")
```
#note: we need to find a way to get the locations for these hospitals. Currently, we only have the names of the hospitals, which is quite problematic.

```{r}
hospitals_final %>% 
mutate(price_level=case_when(billing_amount<13507 ~ "low", 
                                  billing_amount>=13507 & billing_amount<=25258 ~ "normal", 
                                  billing_amount>25258 & billing_amount<=37734 ~ "high", 
                                  billing_amount>37734 ~ "very high")) %>% 
  count(insurance_provider, price_level) %>% 
  ggplot(aes(x=insurance_provider, y=n, fill=price_level))+
  geom_col(position="dodge")
```

```{r}
hospitals_final %>% 
mutate(price_level=case_when(billing_amount<13507 ~ "low", 
                                  billing_amount>=13507 & billing_amount<=25258 ~ "normal", 
                                  billing_amount>25258 & billing_amount<=37734 ~ "high", 
                                  billing_amount>37734 ~ "very high")) %>% 
  count(medical_condition, price_level) %>% 
  ggplot(aes(x=medical_condition, y=n, fill=price_level))+
  geom_col(position = "dodge")
```
#1. Need to find a way to convert dates into numeric value for day of the year. 
#2. Evaluate hospital quality (length of stay times billing amount). Lower value is higher quality. 
#3. Group by medical condition and give top 5 hospitals.

```{r}
library(lubridate)
```

```{r}
hospitals_final %>%
  mutate(discharge_date = as.Date(discharge_date), 
         date_of_admission = as.Date(date_of_admission),
         time_spent = discharge_date - date_of_admission) %>%
  filter(!is.na(time_spent)) %>% 
  separate(time_spent, into="length_of_stay", sep = " ") %>% 
  mutate(length_of_stay=as.numeric(length_of_stay))
```

#can use this code to find top 5 hospitals for each condition with slight adjustment
```{r}
hospitals_final %>%
  mutate(discharge_date = as.Date(discharge_date), 
         date_of_admission = as.Date(date_of_admission),
         time_spent = discharge_date - date_of_admission) %>%
  filter(!is.na(time_spent)) %>% 
  separate(time_spent, into="length_of_stay", sep = " ") %>% 
  mutate(length_of_stay=as.numeric(length_of_stay)) %>% 
  mutate(hospital_quality=log10(length_of_stay*billing_amount)) %>% 
  arrange(hospital_quality) %>% 
  filter(medical_condition=="Cancer") %>% 
  select(hospital, hospital_quality, admission_type, medical_condition) %>% 
  head(5)
```
```{r}
table(hospitals_final$medical_condition)
```



```{r}
hospitals_final %>%
  mutate(discharge_date = as.Date(discharge_date), 
         date_of_admission = as.Date(date_of_admission),
         time_spent = discharge_date - date_of_admission) %>%
  filter(!is.na(time_spent)) %>% 
  separate(time_spent, into="length_of_stay", sep = " ") %>% 
  mutate(length_of_stay=as.numeric(length_of_stay)) %>% 
  mutate(hospital_quality=log10(length_of_stay*billing_amount)) %>% 
  arrange(hospital_quality) %>% 
  filter(medical_condition=="Arthritis") %>% 
  select(hospital, hospital_quality, admission_type, medical_condition) %>% 
  head(5)
```
```{r}
hospitals_final %>%
  mutate(discharge_date = as.Date(discharge_date), 
         date_of_admission = as.Date(date_of_admission),
         time_spent = discharge_date - date_of_admission) %>%
  filter(!is.na(time_spent)) %>% 
  separate(time_spent, into="length_of_stay", sep = " ") %>% 
  mutate(length_of_stay=as.numeric(length_of_stay)) %>% 
  mutate(hospital_quality=log10(length_of_stay*billing_amount)) %>% 
  arrange(hospital_quality) %>% 
  filter(medical_condition=="Asthma") %>% 
  select(hospital, hospital_quality, admission_type, medical_condition) %>% 
  head(5)
```
```{r}
hospitals_final %>%
  mutate(discharge_date = as.Date(discharge_date), 
         date_of_admission = as.Date(date_of_admission),
         time_spent = discharge_date - date_of_admission) %>%
  filter(!is.na(time_spent)) %>% 
  separate(time_spent, into="length_of_stay", sep = " ") %>% 
  mutate(length_of_stay=as.numeric(length_of_stay)) %>% 
  mutate(hospital_quality=log10(length_of_stay*billing_amount)) %>% 
  arrange(hospital_quality) %>% 
  filter(medical_condition=="Diabetes") %>% 
  select(hospital, hospital_quality, admission_type, medical_condition) %>% 
  head(5)

```
```{r}
hospitals_final %>%
  mutate(discharge_date = as.Date(discharge_date), 
         date_of_admission = as.Date(date_of_admission),
         time_spent = discharge_date - date_of_admission) %>%
  filter(!is.na(time_spent)) %>% 
  separate(time_spent, into="length_of_stay", sep = " ") %>% 
  mutate(length_of_stay=as.numeric(length_of_stay)) %>% 
  mutate(hospital_quality=log10(length_of_stay*billing_amount)) %>% 
  arrange(hospital_quality) %>% 
  filter(medical_condition=="Hypertension") %>% 
  select(hospital, hospital_quality, admission_type, medical_condition) %>% 
  head(5)
```



```{r}
obesity <- hospitals_final %>%
  mutate(discharge_date = as.Date(discharge_date), 
         date_of_admission = as.Date(date_of_admission),
         time_spent = discharge_date - date_of_admission) %>%
  filter(!is.na(time_spent)) %>% 
  separate(time_spent, into="length_of_stay", sep = " ") %>% 
  mutate(length_of_stay=as.numeric(length_of_stay)) %>% 
  mutate(hospital_quality=log10(length_of_stay*billing_amount)) %>% 
  arrange(hospital_quality) %>% 
  filter(medical_condition=="Obesity") %>% 
  select(hospital, hospital_quality, admission_type, medical_condition) %>% 
  head(5)
```



```{r}

ui <- dashboardPage(
  dashboardHeader(title = "Best Hospitals for Care"),
  dashboardSidebar(disable = TRUE),
  dashboardBody(
    fluidRow(
      box(
        title = "Table Settings", width = 3,
        selectInput("medical_condition", "Select Condition", choices = c("Arthritis", "Asthma", "Cancer", "Diabetes", "Hypertension", "Obesity"), 
                    selected = "Cancer"),
        selectInput("insurance_provider", "Select Insurance Provider", choices = c("Aetna", "Blue Cross", "Cigna", "Medicare", "UnitedHealthcare"),
                    selected = "Medicare")
      ),
      box(
        title = "Top 5 Hospitals",
        width = 7,
        dataTableOutput("table")
      )
    )
  )
)

server <- function(input, output, session) {
  
  session$onSessionEnded(stopApp)
  
  # Define top5 as a reactive expression
  top5 <- reactive({
    hospitals_final %>%
      mutate(discharge_date = as.Date(discharge_date), 
             date_of_admission = as.Date(date_of_admission),
             time_spent = discharge_date - date_of_admission) %>%
      filter(!is.na(time_spent)) %>% 
      separate(time_spent, into = "length_of_stay", sep = " ") %>% 
      mutate(length_of_stay = as.numeric(length_of_stay)) %>% 
      mutate(hospital_quality = log10(length_of_stay * billing_amount))
  })
  
  output$table <- renderDataTable({
    data <- top5()
    
    # Filter the data based on selected medical_condition
    data <- data[data$medical_condition == input$medical_condition,]
    
    # Filter the data based on selected insurance provider
    data <- data[data$insurance_provider == input$insurance_provider,]
    
    # Select top five hospitals based on hospital_quality
    data <- data %>% 
      arrange(hospital_quality) %>% 
      head(5) %>%
      select(hospital, hospital_quality, admission_type, medical_condition)
    
    # Return the filtered data for rendering in the DataTable
    data
  })
}

shinyApp(ui, server)

```





\\ Part 4


```{r}
hospitals_location <- read_csv("data/hospital_locations.csv") %>%
  clean_names()
```

```{r}
hospitals_location <- hospitals_location[, c("index", "id", "name", "address", "city", "state", "zip", "type", "status", "latitude", "longitude", "website")] %>%
  mutate(website = na_if(website, "NOT AVAILABLE"))
```





```{r}
x <- hospitals_final %>%
  mutate(discharge_date = as.Date(discharge_date), 
         date_of_admission = as.Date(date_of_admission),
         time_spent = discharge_date - date_of_admission) %>%
  filter(!is.na(time_spent)) %>% 
  separate(time_spent, into="length_of_stay", sep = " ") %>% 
  mutate(length_of_stay=as.numeric(length_of_stay)) %>% 
  mutate(hospital_quality=log10(length_of_stay*billing_amount)) %>% 
  arrange(hospital_quality) %>% 
  filter(medical_condition=="Hypertension") %>% 
  select(hospital, hospital_quality, admission_type, medical_condition) %>% 
  head(5)
```

```{r}
for(i in 1:nrow(x)) {
    row <- x[i,]
    print(strsplit(row$hospital, " ")[1][[1]][1])
}
```



```{r}
for(i in 1:nrow(x)) {
    row <- x[i,]
    w <- strsplit(row$hospital, " ")[[1]][1]
    w <- str_replace_all(w, "[^[:alnum:]]", " ")
    print(hospitals_location[grep(w, tolower(hospitals_location$name), ignore.case = TRUE), c("name", "zip", "latitude", "longitude")])
}
```


```{r}
combined_results <- list()

for(i in 1:nrow(x)) {
    row <- x[i,]
    w <- strsplit(row$hospital, " ")[[1]][1]
    w <- str_replace_all(w, "[^[:alnum:]]", " ")
    # Append each subset to the list
    combined_results[[i]] <- hospitals_location[grep(w, tolower(hospitals_location$name), ignore.case = TRUE), c("name", "zip", "latitude", "longitude")]
}

# Combine all the subsets into a single dataframe
result <- do.call(rbind, combined_results)

result
```


```{r}
result1 <- hospitals_final %>%
  mutate(discharge_date = as.Date(discharge_date), 
         date_of_admission = as.Date(date_of_admission),
         time_spent = discharge_date - date_of_admission) %>%
  filter(!is.na(time_spent)) %>% 
  separate(time_spent, into="length_of_stay", sep = " ") %>% 
  mutate(length_of_stay=as.numeric(length_of_stay)) %>% 
  mutate(hospital_quality=log10(length_of_stay*billing_amount)) %>%
  arrange(hospital_quality) %>% 
  filter(medical_condition=="Hypertension") %>% 
  select(hospital, hospital_quality, admission_type, medical_condition) %>% 
  head(5) %>%
  rowwise() %>%
  mutate(
    w = strsplit(hospital, " ")[[1]][1],
    w = str_replace_all(w, "[^[:alnum:]]", " "),
    combined_results = list(hospitals_location[grep(w, tolower(hospitals_location$name), ignore.case = TRUE), c("name", "zip", "latitude", "longitude")])
  ) %>%
  pull(combined_results) %>%
  do.call(rbind, .)
result1
```

\\ App Map

```{r}
ui <- dashboardPage(
  dashboardHeader(title = "Best Hospitals for Care"),
  dashboardSidebar(disable = TRUE),
  dashboardBody(
    fluidRow(
      box(
        title = "Table Settings", width = 3,
        selectInput("medical_condition", "Select Condition", choices = c("Arthritis", "Asthma", "Cancer", "Diabetes", "Hypertension", "Obesity"), 
                    selected = "Cancer"),
        selectInput("insurance_provider", "Select Insurance Provider", choices = c("Aetna", "Blue Cross", "Cigna", "Medicare", "UnitedHealthcare"),
                    selected = "Medicare")
      ),
      box(
        title = "Top 5 Hospitals",
        width = 7,
        dataTableOutput("table")
      )
    ),
    fluidRow(
      box(
        title = "Matched Hospitals",
        width = 12,
        dataTableOutput("matched_hospitals_table")
      )
    )
  )
)

# Define server logic
server <- function(input, output, session) {
  
  session$onSessionEnded(stopApp)
  
  # Define top5 as a reactive expression
  top5 <- reactive({
    hospitals_final %>%
      mutate(discharge_date = as.Date(discharge_date), 
             date_of_admission = as.Date(date_of_admission),
             time_spent = discharge_date - date_of_admission) %>%
      filter(!is.na(time_spent)) %>% 
      separate(time_spent, into = "length_of_stay", sep = " ") %>% 
      mutate(length_of_stay = as.numeric(length_of_stay)) %>% 
      mutate(hospital_quality = log10(length_of_stay * billing_amount)) %>%
      arrange(hospital_quality) %>% 
      filter(medical_condition == input$medical_condition) %>% 
      head(5) %>%
      select(hospital, hospital_quality, admission_type, medical_condition)
  })
  
  output$table <- renderDataTable({
    data <- top5()
    
    # Return the filtered data for rendering in the DataTable
    data
  })
  
  # Define the reactive function to find matches from the other dataset
  matched_hospitals <- reactive({
    x <- top5()  # Use the same filtering logic as top5 reactive
    
    combined_results <- list()
    
    # Loop through each hospital in the filtered dataset
    for(i in 1:nrow(x)) {
      row <- x[i,]
      w <- strsplit(row$hospital, " ")[[1]][1]
      w <- str_replace_all(w, "[^[:alnum:]]", " ")
      # Append each subset to the list
      combined_results[[i]] <- hospitals_location[grep(w, tolower(hospitals_location$name), ignore.case = TRUE), c("name", "zip", "latitude", "longitude")]
    }
    
    # Combine all the subsets into a single dataframe
    result <- do.call(rbind, combined_results)
    
    result
  })
  
  # Render the matched hospitals datatable
  output$matched_hospitals_table <- renderDataTable({
    matched_hospitals()
  })
}

# Run the application
shinyApp(ui, server)
```



```{r message=FALSE, warning=FALSE}
library(ggmap)
``` 

```{r}
register_stadiamaps("d67f62a2-bd90-430c-a40a-9ad6e095b888", write = FALSE)
```

```{r}
result %>%
  select(longitude, latitude) %>%
  summary()
```

```{r}
lat <- c(29.75  , 35.70  )
long <- c(-94.95, -80.71)
bbox <- make_bbox(long, lat, f=0.03)
```

```{r}
map <- get_stadiamap(bbox, maptype = "stamen_terrain", zoom=7)
```

```{r}
ggmap(map)
```

```{r}
ggmap(map) +
  geom_point(data=result, aes(longitude, latitude), size = 2)+
  labs(x = "Longitude", 
       y = "Latitude",
       title = "Best hospitals")
```


\\ App MAP.2

```{r}
library(sf)
```



```{r}
ui <- dashboardPage(
  dashboardHeader(title = "Best Hospitals for Care"),
  dashboardSidebar(disable = TRUE),
  dashboardBody(
    fluidRow(
      box(
        title = "Table Settings", width = 3,
        selectInput("medical_condition", "Select Condition", choices = c("Arthritis", "Asthma", "Cancer", "Diabetes", "Hypertension", "Obesity"), 
                    selected = "Cancer"),
        selectInput("insurance_provider", "Select Insurance Provider", choices = c("Aetna", "Blue Cross", "Cigna", "Medicare", "UnitedHealthcare"),
                    selected = "Medicare")
      ),
      box(
        title = "Top 5 Hospitals",
        width = 7,
        dataTableOutput("table")
      )
    ),
    fluidRow(
      box(
        title = "Matched Hospitals",
        width = 6,
        dataTableOutput("matched_hospitals_table")
      ),
      box(
        title = "Map of Matched Hospitals",
        width = 6,
        plotOutput("map_plot")
      )
    )
  )
)

# Define server logic
server <- function(input, output, session) {
  
  session$onSessionEnded(stopApp)
  
  # Define top5 as a reactive expression
  top5 <- reactive({
    hospitals_final %>%
      mutate(discharge_date = as.Date(discharge_date), 
             date_of_admission = as.Date(date_of_admission),
             time_spent = discharge_date - date_of_admission) %>%
      filter(!is.na(time_spent)) %>% 
      separate(time_spent, into = "length_of_stay", sep = " ") %>% 
      mutate(length_of_stay = as.numeric(length_of_stay)) %>% 
      mutate(hospital_quality = log10(length_of_stay * billing_amount))
  })
  
  output$table <- renderDataTable({
    data <- top5()
    
    # Filter the data based on selected medical_condition
    data <- data[data$medical_condition == input$medical_condition,]
    
    # Filter the data based on selected insurance provider
    data <- data[data$insurance_provider == input$insurance_provider,]
    
    # Select top five hospitals based on hospital_quality
    data <- data %>% 
      arrange(hospital_quality) %>% 
      head(5) %>%
      select(hospital, hospital_quality, admission_type, medical_condition)
    
    # Return the filtered data for rendering in the DataTable
    data
  })
  
# Define the reactive expression to find matches from the other dataset
matched_hospitals <- reactive({
  data <- top5()  # Use the same filtering logic as top5 reactive
  
  combined_results <- list()
  
  # Loop through each hospital in the filtered dataset
  for(i in 1:nrow(data)) {
    row <- data[i,]
    w <- strsplit(row$hospital, " ")[[1]][1]
    w <- str_replace_all(w, "[^[:alnum:]]", " ")
    # Append each subset to the list
    combined_results[[i]] <- hospitals_location[grep(w, tolower(hospitals_location$name), ignore.case = TRUE), c("name", "zip", "latitude", "longitude")]
  }
  
  # Combine all the subsets into a single dataframe
  result <- do.call(rbind, combined_results)
  
  result
})

# Render the map of matched hospitals
output$map_plot <- renderPlot({
  result <- matched_hospitals()
  
  # Get the bounding box for the map
  lat <- c(summary(result$latitude)["Min."], summary(result$latitude)["Max."])
  long <- c(summary(result$longitude)["Min."], summary(result$longitude)["Max."])
  bbox <- make_bbox(long, lat, f = 0.03)
  
  # Get the map
  map <- get_stamenmap(bbox, maptype = "terrain", zoom = 7)
  
  # Plot the map
  ggmap(map) +
    geom_point(data = result, aes(longitude, latitude), size = 2) +
    labs(x = "Longitude", 
         y = "Latitude",
         title = "Best Hospitals")
})

}

# Run the application
shinyApp(ui, server)

```






